COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-11 paper/bare_conf.tex
Added new plot. Minor refactor

OP2 adopts a programming model in which computations are expressed through self-contained functions, referred to as ``kernels''. A kernel is applied to all elements in a set of mesh components, such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.

@@

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value, which prevents the adoption of standard techniques for cache locality, further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-results}, in which manual BLAS-based implementations are evaluated, the small dimension of the involved matrices and loss in data locality can limit the performance gain, if any.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh with Lagrange $p=1$ elements.}
\label{code:helmholtz}
\end{algorithm}

\begin{algorithm}[t]
\small
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh with Lagrange $p=1$ elements.}
\label{code:burgers}
\end{algorithm}

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Data alignment and padding are for an AVX architecture. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
\label{code:helmholtz-licm}
\end{algorithm}

Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// Call $Load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, \emph{licm} and \emph{op-vect} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}. Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length. The original size of the $j$-$k$ iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
%~~// Declaration of basis function matrices \\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when \emph{split} is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, the high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

@@

Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} (other optimization levels perfomed, in general, slightly worse) and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, the optimization level \texttt{-O3} was used. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

>>>>

OP2 adopts a programming model in which computations are expressed through self-contained functions, referred to as ``kernels''. A kernel is applied to all elements in a set of mesh components, such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{pyop2isc},~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem in the context of the finite element method.

@@

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value, which prevents the adoption of standard techniques for cache locality, further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the involved matrices and the potential loss in data locality can limit the performance gain, if any.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh using Lagrange $p=1$ elements.}
\label{code:helmholtz}
\end{algorithm}

\begin{algorithm}[t]
\small
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
\label{code:burgers}
\end{algorithm}

@@

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable performance improvements. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4] ALIGN;\\
~~~~double LI$\_$1[4] ALIGN;\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz} for an AVX architecture. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in the $r$ loop.}
\label{code:helmholtz-licm}
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
~~~~// Do loop-invariant code motion \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~~~// $load$ and $set$ intrinsics \\
~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
~~~~~~~~// ...\\
~~~~~~$\rbrace$\\
~~~~// Remainder loop (from $j=4$ to $j=6$)\\
~~$\rbrace$\\
~~// Restore the storage layout:\\
~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
~~~~$\rbrace$\\
~~$\rbrace$\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem when \emph{op-vect} is applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}. Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length. The original size of the $j$-$k$ iteration space (i.e. before padding was applied) was 6$\times$6. In this example, the unroll-and-jam factor is 1.}
\label{code:helmholtz-opvect}
\end{algorithm}

@@

\begin{algorithm}[t]
\small
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
%~~// Declaration of basis function matrices \\
%~~\\
%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
%~~~~double LI$\_$0[4] ALIGN;\\
%~~~~double LI$\_$1[4] ALIGN;\\
%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
%~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
$\rbrace$
\caption{Local assembly code generated by Firedrake for the Helmholtz problem when \emph{split} is applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
\label{code:helmholtz-split}
\end{algorithm}

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, the high-level specification is translated by the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

@@

Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, the optimization level \texttt{-O3} was used. Other optimization levels perfomed, in general, slightly worse. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.

These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method, whereas the element family, Lagrange, is fixed. Intuitively, the bigger the element shape and $p$, the larger is the iteration space. Triangles, tetrahedron, and prisms were tested as element shape. For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three element shapes is, respectively, $3$, $4$, and $6$. Moving to bigger shapes has the effect of increasing the number of basis function arrays, since, intuitively, the behaviour of the equation has now to be approximated also along a third axis. On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays). A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations. In such a large space of problems, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
